---
title: "Phenology model workflow"
author: "C. Susannah Tysor"
date: '2020-08-25'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(zoo)
library(flowers)
library(dplyr)

source('phenology_functions.R')
```

# Conceptual Analysis

I have a record of days that lodgepole pine were recorded flowering or not flowering during the general flowering period. Trees were not observed every day and the start or end of the flowering period may have been missed.

The flowering observations ($y$) were simplified to first recorded (began flowering) and last recorded (finished flowering) - discrete events. The amount of forcing accumulated at each event day was calculated using Risto Sarvas's 1972 curve.

I want to know how much forcing after January 1 must accumulate for trees to achieve each event.

$y_i \sim \mathcal{N}(f_i, \sigma)$

I wish to account for several kinds of structure in the data:

- year of observation
- site where tree was grown
- provenance the tree's parents were sourced from
- clone identity (multiple ramets of clones are grown at seed orchard sites)

So

$f_i = \mu + alpha_{site,i} + \alpha_{prov,i} + \alpha_{year,i} + \alpha_{clone,i}$

where mu is the average amount of forcing it takes for a tree to reach an event and the $\alpha$s describe how each group in each cluster (site, prov, year, clone) effects the overall amount of forcing required to reach the event.

I want priors on parameters to be regularizing and weakly informative.

# Define observational space

There are several thousand observations of each event. Each observation is the amount of forcing units accumulated when the event occurs for a given tree. 



So, how much forcing is accumulated over the year?

```{r}

# forcing units for 15 years and at sites included in this study
forcing <- read.csv('../phenolology/data/all_clim_PCIC.csv', header=TRUE) %>%
  dplyr::filter(forcing_type=="ristos")

ggplot(forcing, aes(x=DoY, y=sum_forcing, colour = interaction(Site,Year))) +
  geom_line() +
  theme(legend.position = "none") +
  ggtitle("forcing (ristos) accumulated at Day of Year")



```

What are reasonable amounts of forcing? Lodgepole in BC flower in late spring.

```{r}
# likely Day of Year limits for flowering
early_limit <- as.numeric(strftime("2020-04-15", format = "%j"))
late_limit <- as.numeric(strftime("2020-06-30", format = "%j"))

ff <- dplyr::filter(forcing, DoY > early_limit & DoY < late_limit)

ggplot(ff, aes(x=sum_forcing)) +
         geom_histogram(bins=60) +
  ggtitle("Range of forcing accummulations Apr 15 - June 30")

ggplot(ff, aes(x=DoY, y=sum_forcing, colour = interaction(Site,Year))) +
  geom_line() +
  theme(legend.position = "none") +
  ggtitle("forcing (ristos) accumulated at Day of Year", subtitle = "April 15 - June 30")
```


```{r}
# May and June alone
apr15 <- as.numeric(strftime("2020-04-15", format = "%j"))
may31 <- as.numeric(strftime("2020-05-31", format = "%j"))

may15 <- as.numeric(strftime("2020-05-15", format = "%j"))
jun30 <- as.numeric(strftime("2020-06-30", format = "%j"))

ffstart <- dplyr::filter(forcing, DoY > apr15 & DoY < may31)
range(ffstart$sum_forcing)

ffend <- dplyr::filter(forcing, DoY > may15 & DoY < jun30)
range(ffend$sum_forcing)

```
Flowering is more likely to start in the first part (`r range(ffstart$sum_forcing)` forcing units) and end in the second part (`r range(ffend$sum_forcing)` forcing units) of this time period.

Each event is associated with a year, provenance, site, and clone. There are 14 years, 7 sites and provenances, and several hundred clones.


# Construct summary statistics

Histograms of $y_i$ and each $\alpha_{cluster}$.

# Model development

$y_i$ are the observations of accumulated forcing. They are positive and continuous and should be between about 200 and 1000. 

$y_i$'s value can be modeled as coming from a distribution with a mean forcing accumulation ($f_i$) and standard deviation $\sigma$ that represents our uncertainty in the observation - how good our temperature measurements are, how good the forcing unit accumulation function is, and whether the event was observed on the correct day. 

## sigma

I think that most error is probably within about 1 day of forcing units in the observation period.

```{r}

# get 1 day forcing accumulations

janjun <- forcing %>%
  filter(DoY < late_limit) 

ggplot(janjun, aes(x=forcing)) +
  geom_histogram() +
  ggtitle("Forcing accumulation over 1 day periods in May and June") +
  geom_vline(xintercept = mean(janjun$forcing))

# 90th percentile of forcing diffs
dextreme <- quantile(janjun$forcing, 1)

pnorm(dextreme, 0, sd=6) # sd to cap at extreme
pexp(6, 1)
pnorm(6, 0, 2)

hist(rnorm(1000, 0, sd=6), breaks=30, xlim = c(-25, 25)) 
abline(v = dextreme)
hist(rexp(1000, 1), breaks=30)
abline(v=6)
hist(abs(rnorm(1e3, 0, 2)), breaks=30) 
abline(v=6)

```

So,

$y_i ~ \mathcal{N}(f_i, \sigma)$

$\sigma \sim \mathcal{Exp}(1)$

$f_i$ is a function of $\mu$ and the $\alpha$ effects. 

## mu

$\mu$ represents the population average for forcing required to achieve the flowering event of interest. It must be positive and continuous. Values less than 75 and greater than ~1000 would be quite unusual for any phase.

For starting, values less than ~75 and greater than ~600 would be surprising.
For finishing, values less than ~160 and greater than ~950 would be surprising.

### mu start
For start events



```{r}
pnorm(600, mean = (75+600)/2, sd = 50) #start

hist(rnorm(1000, 335, 50), breaks=60)
abline(v=600)
```
$\mu \sim \mathcal{N}(335,50)$

### mu end

For end events:

$\mu \sim \mathcal{N}(555,50)$

```{r}
pnorm(950, mean = (160+950)/2, sd = 80) #end
hist(rnorm(1000, 555, 80), breaks=60)
```
## alpha 

The $\alpha$ effects can be modeled as changes from $\mu$. I believe (hope?) they're zero, but they *could* be as much as a week's worth of forcing.

### alpha mean

A flowering period of 2 weeks for a population would be very long and a flowering period of a week very long for an individual tree, so an effect ($\alpha$) that changed the timing of flowering by even a few days would be meaningful. An alpha that shifted flowering by 1 weeks would be quite extreme. How much forcing can accumulate in 1 week prior to an event?

[Since there aren't extreme variations in the length of flowering period, using the first part of the flowering period might be a better guide for this. Currently using January - June]

```{r}


week_forcing <- janjun %>%
  group_by(Site, Year) %>%
  summarise(sum_forcing_week = zoo::rollsum(forcing, 7))

ggplot(week_forcing, aes(x=sum_forcing_week)) +
  geom_histogram() +
  ggtitle("Forcing accumulation over 1 week periods from January 1 to June 30") +
  geom_vline(xintercept = mean(week_forcing$sum_forcing_week))

# 90th percentile of forcing that can accumulate in a week
asextreme <- quantile(week_forcing$sum_forcing_week, .90) # because forcing accumulates so much faster at the end of june when flowering is quite unlikely, capping at 90%

pnorm(asextreme, mean=0, sd=14) # what sd caps at extreme forcing accumulation

```
`r asextreme` would be a very extreme alpha.


$\alpha_{cluster} \sim \mathcal{N}(\mu_{cluster},\sigma_{cluster})$

$\mu_{cluster} \sim \mathcal{N}(0, 14)$

```{r}
hist(rnorm(1000, mean=0, sd=14), breaks=50) 
abline(v=asextreme)
```


### alpha sigmas

$\sigma_{cluster}$ describes variation between clusters. I think a very extreme value for this would be a difference of a week, maxing out at `r asextreme` forcing units. As calculated above, this requires a sigma of ~14.

For clone, which has hundreds of clusters, we can use an exponential

$\sigma_{cluster} \sim \mathcal{Exp}(0.7)$

```{r}
pexp(14, rate = 0.5)
hist(rexp(1000, 0.5), breaks=40)
abline(v=14)
```

For site and clone (7 clusters), and maybe year (14 clusters), we should use something with thinner tails because we don't have many clusters to estimate sd with here.
 
$\sigma_{cluster} \sim \mathcal{HalfNormal}(0, 4)$

```{r}
pnorm(14, 0, 5)
hist(abs(rnorm(1000, 0, 5)), breaks=40)
abline(v=14)
```
 
## Centering

Factors that have a lot of levels all with limited data, like clone, will probably sample best if non-centered.

Some of my factors are very unbalanced with some levels very well represented and others not.

```{r}
phenbe <- filter_start_end() 

phen_by_sex_and_side <- function(df, sex, side) {
  outdf <- select_data(df, sex, side)
  outdf$Side <- side
  outdf$Sex <- sex
  return(outdf)
}

male_begin <- phen_by_sex_and_side(phenbe, "MALE", "begin")
male_end <- phen_by_sex_and_side(phenbe, "MALE", "end")
female_begin <- phen_by_sex_and_side(phenbe, "FEMALE", "begin")
female_end <- phen_by_sex_and_side(phenbe, "FEMALE", "end")

phenbe <- rbind(male_begin, male_end, female_begin, female_end)

#   dplyr::select(Year, Site, Provenance, Clone, Side, Sex) %>%
#   mutate_all(as.character) %>%
#   tidyr::pivot_longer( cols=c("Year", "Site", "Provenance", "Clone"), names_to = "Factor", values_to = "Level")


ggplot(phenbe, aes(x=Site)) +
  geom_histogram(stat="count") +
  facet_grid(Sex ~ Side) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

ggplot(phenbe, aes(x=Provenance)) +
  geom_histogram(stat="count") +
  facet_grid(Sex ~ Side) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

ggplot(phenbe, aes(x=Year)) +
  geom_histogram(stat="count") +
  facet_grid(Sex ~ Side) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

ggplot(phenbe, aes(x=Clone)) +
  geom_histogram(stat="count") +
  facet_grid(Sex ~ Side) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 


```

Looks like Clone, Orchard, and Site all need to have non-centering *within* factors. 

- Site threshold: 250
- Provenance threshold: 150
- Clone threshold: 10

Year *might* benefit from non-centering with a threshold around either 150 or 100
