---
title: "Model methods"
author: "C. Susannah Tysor"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}

library(flowers)
library(dplyr)
library(brms)
library(ggplot2)
library(tidyr)
library(tidybayes)
library(forcats)
library(ggbeeswarm)
library(lubridate)

knitr::opts_chunk$set(echo = FALSE)
theme_set(theme_dark())

source('phenology_functions.R')
```

## Data

```{r climdat}
siteclim <- read.csv("../lodgepole_climate/processed/PCIC_all_seed_orchard_sites_adjusted.csv")
histclim <- read.csv("data/all_clim_PCIC.csv") %>% # site clim with forcing
   filter(forcing_type == "gdd")  
provclim <- read.csv("../phd/data/OrchardInfo/lodgepole_SPU_climsum.csv")
futclim <- read.csv("../lodgepole_climate/processed/future_daily_temps.csv")
```

Phenology data is available in the `flowers` R package published on Github [TO BE ARCHIVED ON ZENODO]

```{r data}
phendat <- flowers::lodgepole_phenology_event %>%
  mutate(Tree = paste0(Orchard, Clone, X, Y)) # create a unique Tree identifier since original data doesn't always have one
spudat <- read.csv("../phd/data/OrchardInfo/LodgepoleSPUs.csv", header = TRUE, stringsAsFactors = FALSE) 

phenf <- prepare_data(phendat, clim = histclim, spu = spudat)

ggplot(phenf, aes(x = sum_forcing, color = Event_Label, linetype = Sex)) +
  stat_ecdf() +
  labs(title = "Cumulative distribution of accumulated forcing for flowering events", caption = "raw data") +
  scale_colour_viridis_d() +
  theme_dark(base_size = 18) +
  ylab("") +
  xlab("GDD")

# create 4 datasets for 4 models

fbdat <- filter_sex_event(sex = "FEMALE", event = "begin", phenf)
fedat <- filter_sex_event(sex = "FEMALE", event = "end", phenf)

mbdat <- filter_sex_event(sex = "MALE", event = "begin", phenf)
medat <- filter_sex_event(sex = "MALE", event = "end", phenf)



```



```{r ordering}
# order provenances and sites from warmest to coldest
siteMAT <- siteclim %>% 
  mutate(Year = lubridate::year(Date)) %>%
  group_by(Site) %>%
  summarise(MAT = mean(mean_temp_corrected)) %>%
  arrange(MAT)

sitefactororder <- siteMAT$Site

provMAT <- provclim %>% select(SPU_Number, MAT) %>%
  full_join(spudat) %>%
  filter(SPU_Name %in% unique(phenf$Provenance)) %>%
  select(MAT, SPU_Name) %>%
  distinct() %>%
  arrange(MAT)

provfactororder <- provMAT$SPU_Name %>% 
  stringr::str_replace_all(pattern = "\\s", replacement = "\\.")  # format names to work with stan output

yearMAT <- siteclim %>% 
  mutate(Year = lubridate::year(Date)) %>%
  right_join(data.frame(Year = as.numeric(unique(phenf$Year)))) %>%
  group_by(Year) %>%
  summarise(MAT = mean(mean_temp_corrected)) %>%
  arrange(MAT)

yearfactororder <- yearMAT$Year

```
## censorship
Make censorship tables (proportion?)
```{r}

censdf <- filter(phenf, Event_Obs %in% c(2,3)) %>%
  group_by(Event_Label, Sex) %>%
  mutate(ng = n()) %>%
  group_by(Event_Label, Sex, censored) %>%
  summarise(prop_cens = n()/unique(ng)) %>%
  arrange(Event_Label, Sex)

ggplot(censdf, aes(x = Sex, y = prop_cens, fill = censored)) +
  geom_bar(stat = "identity") +
  facet_wrap("Event_Label") 

```

## Model

We built separate hierarchical Bayesian intercept-only factor models of the accumulated forcing at each of the four flowering events of interest: start of receptivity (female begin), end of receptivity (female end), start of pollen shed (male begin), and end of pollen shed (male end).

Ideally, the likelihood of begin and end of flowering observations $y_i$ would

$y_i \sim \mathrm{Normal}(\phi_i, \sigma)$

However, interval and end censoring
The likelihood is modified for the 37 to 65% of end censored observations. 

Flowering start data is left censored when a tree is already flowering when it is first observed, so the likelihood is calculated with a cumulative normal distribution function.

$\mathrm{y_i} \sim \mathrm{Cumulative Normal}(\phi_i,\sigma)$

Flowering end data is right censored when a tree is never observed past flowering and a complementary cumulative normal distribution function is used for the likelihood. 

$\mathrm{y_i} \sim \mathrm{Complementary Cumulative Normal}(\phi_i, \sigma)$.

Mean $\phi_i$ is the sum of the overall population mean $\mu$ and offsets $\delta$ for each level of factors Site, Provenance, Year, and Clone.

$\mathrm{\phi_i} = \mu + \delta_{Site, i} + \delta_{Provenance, i} + \delta_{Year, i} + \delta_{Clone, i}$

Priors limit parameter estimates to forcing accumulations possible at our sites during late spring, when lodgepole pine are known to flower in British Columbia (@owensReproductiveBiologyLodgepole2006). Offset standard deviations $\sigma_{factor}$ were roughly constrained by both estimates of possible forcing accumulation and the approximately known length of the flowering period. Possible forcing accumulation ranges were calculated using daily PCIC weather data from 1945-2012 at all of our sites. As Site, Provenance, and Year have relatively few levels with which to estimate $\sigma_{factor}$, half normal priors with their relatively thin tails were used to help constrain the parameter. Additional details on prior choice can be found in the conceptual analysis (supplementary materials).

Priors for the begin models differ for those of the end models as forcing accumulates faster in later spring than early spring and flowering must begin before it can end.

Priors for the start of receptivity and pollen shed models were
$$\begin{align}
\mu & \sim \mathrm{Normal}(400,100) \\
\sigma & \sim \left|\mathrm{Normal}(0,15)\right| \\
\sigma_{factor, j} & \sim \left|\mathrm{Normal}(0,9)\right| \\
\end{align}$$



```{r model}
# This model block is faster if you run models in parallel

# initialize parameter values on the right order of magnitude
initpars <- lapply(1:6, function(id) list(sigma = 30, Intercept = 300))

# model formula
bform <- brmsformula(sum_forcing | cens(censored, upper) ~ 1 + (1|Site) + (1|Provenance) + (1|Clone) + (1|Year) + (1|Tree))

# model prior
bprior <- c(prior("normal(400,100)", class = "Intercept"),
            prior("normal(0,15)", class = "sigma"),
            prior("normal(0,9)", class = "sd"))
niter <- 4000
ncores <- 6
nchains <- 6

fbfit <- brm(bform, data = fbdat,
             save_model = "female_begin.stan",
             file = "female_begin",
             prior = bprior,
             inits = initpars,
             iter = niter,
             cores = ncores,
             chains = nchains,
             sample_prior = TRUE,
             save_pars = save_pars(all = TRUE),
             file_refit = "on_change")
#fbfit <- readRDS("female_begin.rds")

fefit <- brm(bform, data = fedat,
             save_model = "female_end.stan",
             file = "female_end",
             prior = bprior,
             inits = initpars,
             iter = niter,
             cores = ncores,
             chains = nchains,
             sample_prior = TRUE,
             save_pars = save_pars(all = TRUE),
             file_refit = "on_change")
#fefit <- readRDS("female_end.rds")

mbfit <- brm(bform, data = mbdat,
             save_model = "male_begin.stan",
             file = "male_begin",
             prior = bprior,
             inits = initpars,
             iter = niter,
             cores = ncores,
             chains = nchains,
             sample_prior = TRUE,
             save_pars = save_pars(all = TRUE),
             file_refit = "on_change")
#mbfit <- readRDS("male_begin.rds")

mefit <- brm(bform, data = medat,
             save_model = "male_end.stan",
             file = "male_end",
             prior = bprior,
             inits = initpars,
             iter = niter,
             cores = ncores,
             chains = nchains,
             sample_prior = TRUE,
             save_pars = save_pars(all = TRUE),
             file_refit = "on_change")
#mefit <- readRDS("male_end.rds")

modells <- list(fb = readRDS("female_begin.rds"),
                fe = readRDS("female_end.rds"),
                mb = readRDS("male_begin.rds"),
                me = readRDS("male_end.rds"))

```
Models were fit in Stan (`rstan` vers. 2.21.2 [@standevelopmentteam2020]) via the R (R version 4.0.3 (202-10-10)) package `brms` (vers. 2.15.0 [@Burkner2017a]). Stan uses the No-U-Turn Sampler, an efficient Markov Chain Monte Carlo method that extends the Hamiltonian Monte Carlo algorithm [@carpenter2017], to sample the joint posterior. 

### Diagnostics
We ran `r nchains` independent chains of `r niter` iterations each. `r niter/2` warmup iterations were dropped from each chain for `r nchains * (niter/2)` total samples. Model convergence and performance were considered good; $\hat R$ was $< 1.01$ and bulk and tail effective sample size was more than 100x the number of chains for all parameter values in all models [@vehtari2019]. There were no divergences. Visual inspection of energy plots and rank plots showed acceptable sampling behavior.


# Results


```{r samplepost}
nsamp <- 2000
seed <- 1657
```

For efficiency, `r nsamp` posterior samples were used for plotting results.

## Means
```{r means, cache=TRUE}
#vars <- tidybayes::get_variables(fbfit)

labdf <- data.frame(Sex = c("FEMALE", "FEMALE", "MALE", "MALE"), event = c('begin', 'end', 'begin', 'end'), model = c('fb', 'fe', 'mb', 'me'))

means <- purrr::map(modells, gather_means_draws) %>%
  bind_rows(.id = "model") %>%
  left_join(labdf)

ggplot(means, aes(y = fct_rev(event), x = .value, colour = Sex)) +
  stat_halfeye(position = "dodge") +
  scale_colour_viridis_d() +
  labs(title = "Population mean", caption = "2000 draws from the posterior") +
  ylab("") +
  xlab("GDD") +
  theme_dark(base_size = 18) +
  theme(legend.position = "top") +
  scale_x_continuous(breaks = scales::pretty_breaks(n=10))


```
Overall variation in flowering is ~15 forcing units for start and 17 for end. In general, variation is smaller for start events than end events, likely because forcing increases monotonically.

Estimates for variation among Year, Site, and Provenance are relatively uncertain because the number of levels within each of these factors was small (7-15), while there were hundreds of clones and trees.

Within-clone variance (sd Tree) is very low, but higher for males than females. Mean estimates of variation within provenance (among Clone variation) is higher than variation among provenances, but the posterior is quite wide for Provenance and the posterior has a great deal of weight beyond the upper limits of the Clone estimates.

Year explains more of the variation than all other factors except for the female end model where Site is more important.

Female begin: Year, Clone, Site, Provenance, Tree
Male begin: Year, Clone, Provenance, Site, Tree
Female end: Site, Year, Clone, Provenance, Tree
Male end: Year, Site, Clone, Provenance, Tree

Year: me, fe, mb, fb
Site: fe, me, fb, mb
Prov: me, fe, fe, mb
Clone: fe, me, fb, mb
Tree: me, mb, fb, fe

## Variation

```{r variation, cache = TRUE}


variation <- purrr::map(modells, gather_var_draws) %>%
  bind_rows(.id = "model") %>%
  left_join(labdf) %>%
  mutate(.variable = case_when(.variable != "sigma" ~ stringr::str_sub(.variable, 4, -12),
                               .variable == "sigma" ~ "sigma")) %>%
  mutate(.variable = factor(.variable)) %>%
  mutate(.variable = forcats::fct_relevel(.variable, "sigma", "Year", "Site", "Provenance", "Clone", "Tree"))

ggplot(variation, aes(y = fct_rev(.variable), x = .value, colour = .variable, linetype = Sex)) +
  stat_pointinterval(position = "dodge") +
  scale_colour_viridis_d() +
  labs(title = "Standard deviation of \npop mean & offsets", caption = "2000 draws from the posterior") +
  ylab("") +
  xlab("GDD") +
  facet_grid(event ~ .) +
  guides(color = "none", size = "none") +
  theme_dark(base_size = 18) +
  theme(legend.position = "top") 
```
Effects for Tree are generally quite small and are moderate for Provenance. Clone has a wide range of possible effects. PGTIS and PRT have relatively large effects for the end of flowering, tending to stop flowering at a larger forcing amount. (Maybe because of censoring?) Provenances have similar ranked effects across models and events, tending to skew events earlier or later by a small amount of forcing. 

## Offsets
```{r offsets, cache=TRUE}

offsets_raw <- purrr::map(modells, gather_offset_draws) %>%
  bind_rows(.id = "model") %>%
  left_join(labdf) #%>%

# turn brms .variable names into useful names (slow)
varlevel <- offsets_raw$.variable %>% stringr::str_split_fixed("[_\\[\\,]", n=4) %>% data.frame() %>%
  select("X2", "X3")
colnames(varlevel) <- c("factor", "level")

# order factors and factor levels
yearclonetree <- filter(varlevel, factor %in% c("Year", "Clone", "Tree")) %>% distinct()
yctorder <- sort(yearclonetree$level)

spyctorder <- unique(c(sitefactororder, provfactororder, yctorder))

offsets <- offsets_raw %>% cbind(varlevel) %>%
  ungroup() %>%
  mutate(factor = forcats::fct_relevel(factor, "Year", "Site", "Provenance", "Clone", "Tree")) %>%
  mutate(level = forcats::fct_relevel(level, spyctorder))

# slow
offsets_summary <- offsets %>%
  group_by(model, Sex, event, factor, level) %>%
  median_hdci(.value, .width = c(0.5, 0.89)) %>%
  ungroup()


offsets_summary %>%
  select(model, Sex, event, factor, level, .value, .point) %>% distinct() %>%
  ggplot(aes(y=.value, x = model, colour = Sex, shape = event)) +
  geom_quasirandom(alpha = 0.5) +
  facet_wrap("factor") +
  labs(title = "Offset medians", caption = "2000 draws from posterior") +
  geom_hline(yintercept = 0, linetype =3, colour = "darkgray") +
  theme_dark(base_size = 18) +
  ylab("GDD") +
  scale_colour_viridis_d() +
  theme(legend.position = "top") +
  geom_hline(yintercept = 0, linetype = 3)


siter <- filter(offsets, factor == "Site") %>%
  mutate(level = forcats::fct_relevel(level, sitefactororder))

provr <- filter(offsets, factor == "Provenance") %>%
  mutate(level = forcats::fct_relevel(level, provfactororder))

yearr <- filter(offsets, factor == "Year") %>%
  mutate(level = forcats::fct_relevel(level, as.character(yearfactororder)))

ggplot(siter, aes(y=level, x = .value, colour = Sex)) +
  stat_pointinterval() +
  facet_grid(event ~ Sex) +
  ggtitle("Site offsets", subtitle = "Ordered warmest to coldest MAT") +
  theme_dark(base_size = 18) +
  ylab("Site") +
  xlab("GDD") +
  geom_vline(xintercept = 0, linetype =3) +
  xlim(c(-90,90)) +
    scale_colour_viridis_d() +
  theme(legend.position = "top") 

ggplot(provr, aes(y=level, x = .value, colour = Sex)) +
  stat_pointinterval() +
  facet_grid(event ~ Sex) +
  ggtitle("Provenance offsets", subtitle = "Ordered warmest to coldest MAT") +
  theme_dark(base_size = 18) +
  ylab("Provenance") +
  xlab("GDD") +
  geom_vline(xintercept = 0, linetype =3)  +
  xlim(c(-90,90)) +
    scale_colour_viridis_d() +
  theme(legend.position = "top")

ggplot(yearr, aes(y=level, x = .value, colour = Sex)) +
  stat_pointinterval() +
  facet_grid(event ~ Sex) +
  ggtitle("Year offsets", subtitle = "Ordered warmest to coldest MAT") +
  theme_dark(base_size = 18) +
  xlab("GDD") +
  ylab("Year") +
  geom_vline(xintercept = 0, linetype =3, colour = "darkgray")  +
  xlim(c(-90,90)) +
  scale_colour_viridis_d() +
  theme(legend.position = "top")

```

# Retrodictions


## General predictions

```{r general_predictions}


alldatls <- list(fbdat, fedat, mbdat, medat)
alldat <- rbind(fbdat, fedat, mbdat, medat)

allsim <- purrr::map2(alldatls, modells, function(x,y) {simulate_from_model(data = x, model = y, nsamples = 200, seed = seed)}) %>%
  bind_rows()


#may15doy <- lubridate::yday("2021-05-15")

#may15 <- histclim %>% 
 # filter(DoY == may15doy, forcing_type == "gdd") 

ggplot(alldat, aes(x = sum_forcing, y = "observations" , colour = Sex)) +
  stat_dotsinterval( .width = c(0.5, 0.89), point_interval = median_qi) +
  stat_slab(data = allsim, 
                    aes(x = .prediction, y = prediction_type, group=.draw),
                    .width = c(0.5, 0.89), point_interval = median_hdci, 
                    slab_color = "gray65", alpha = 1/10, fill = NA) +
  stat_pointinterval(data = allsim, aes(x = .prediction, y = prediction_type), 
                     .width = c(0.5, 0.89), point_interval = median_hdci ) +
 # stat_dots(data = may15, aes(x = sum_forcing, y = "Forcing at May 15")) +
  theme_bw(base_size = 18) +
  facet_grid(event ~ Sex) +
  labs(title = "Modeled and observed flowering events",  caption = "200 samples from the posterior, 5 for fully crossed predictions") +
  xlab("GDD") +
  ylab("") + 
  scale_colour_viridis_d() +
  theme(legend.position = "none")
#subtitle = "50 & 89% quantile intervals for observations \nand HCDI intervals for model predictions",

```

We used the model to simulate four data sets. First we retrodicted flowering forcing for the observed combinations of trees, clones, years, provenances, and sites (retrodiction - uncensored). Then we censored that data based on the censoring points recorded in our observations (retrodiction - censored). We expect the censored retrodictions to match our observations relatively well. 

We also predicted flowering forcing for a fully crossed version of our original dataset and for new sites, years, provenances, clones, and trees. Since the fully-crossed version of our originally observed data set is 82,768 rows for each sex + event, only 5 samples from the posterior were used to simulate each observation. For new levels, we predicted 1500 forcing observations for all 4 flowering events. We sampled offsets for 5 new sites, years and provenances from the superpopulation posterior distributions for each factor offset ($\mathrm{Normal}(0, \sigma_{factor})$). Each provenance was represented by 10 clones with 2 ramets each.

All simulations used the same 200 draws from the posterior, except for the fully crossed version which used the same 5 which were a subset of the previous 200.

# Add days to preds

See functions in retrodiction functions for this. DoY predictions used only 30 samples from the posterior for efficiency.

```{r predictdaysspecific, fig.height=8}

# first get DoY predictions for specific predictions & retrodictions (uncensored, fully crossed)
# filter simulations for only real sites and years, not new levels and simplify to only columns needed for matching 
specificsim <- filter(allsim, prediction_type %in% c("retrodiction - uncensored", "prediction - full cross")) %>%
  ungroup() %>%
  select(Year, Site, .prediction) %>%
  distinct() 

specific_doy_preds_temp <- forcing_to_doy(filter(histclim, forcing_type == "gdd"), specificsim, aforce = "sum_forcing", bforce = ".prediction", newdoycolname = "newdoycol") 
  
specific_doy_preds <- full_join(filter(allsim, prediction_type %in% c("retrodiction - uncensored", "prediction - full cross")), specific_doy_preds_temp) 
rm(specific_doy_preds_temp)

# get DoY predictions for the general predictions - which means assigning all sites and years to each prediction

# extract heat sum predictions at new sites and years and provs, clones, trees. downsample - 30 draws per "row"/new obs
generalsim <- filter(allsim, prediction_type == "prediction - new levels") %>%
  ungroup() %>%
  select(Sex, event, .row, .draw, .prediction, prediction_type, Site, Year, Provenance, Clone, Tree) %>%
  group_by(Sex, event, .row, prediction_type, Site, Year, Provenance, Clone, Tree) %>%
  slice_sample(n = 30) %>%
  rename(new_Site = Site, new_Year = Year)

# assign each prediction to fully crossed site x climate dataset
climsites <- unique(histclim$Site)
climyears <- unique(histclim$Year)
generalcross <- tidyr::crossing(climsites, climyears, .prediction = unique( generalsim$.prediction) ) %>%
  dplyr::rename(Site = climsites, Year = climyears)

general_doy_preds_temp <- forcing_to_doy(filter(histclim, forcing_type == "gdd"), generalcross, aforce = "sum_forcing", bforce = ".prediction", newdoycolname = "newdoycol")

general_doy_preds <- full_join(generalsim, general_doy_preds_temp)
rm(general_doy_preds_temp)

# general_doy_preds_med <- general_doy_preds %>% group_by(Sex, event, .row, Site, Year) %>%
#   point_interval(prediction, newdoycol, .width = c(0.5, 0.95, .point = median, .interval = hdci))

# so slow
general_doy_preds_med <- general_doy_preds %>% group_by(Sex, event, .row, Site, Year) %>%
  summarise(median = median(newdoycol))

general_doy_preds_med_siteyearsex <- general_doy_preds %>% group_by(Sex, event, Site, Year) %>%
  point_interval(newdoycol, .width = c(0.5, 0.95), .point = median, .interval = hdci)

# time series lines with begin and end for one site a piece on each plot
ggplot(general_doy_preds_med_siteyearsex, aes(x=Year, y = newdoycol, linetype = Sex, colour = event)) +
  #geom_point() +
  geom_line() +
  scale_colour_viridis_d() +
  facet_wrap("Site")


# change doy to dates for plotting
general_doy_preds_med_siteyearsex <- general_doy_preds_med_siteyearsex %>%
  mutate(Date = as.Date("2020-12-31") + newdoycol, .lowerdate = as.Date("2020-12-31") + .lower, .upperdate = as.Date("2020-12-31") + .upper)

# Plot to compare time series
ggplot(general_doy_preds_med_siteyearsex, aes(y = Date, x = Year, colour = Sex, group = Year)) +
  geom_line() +
  facet_grid(Site ~ Sex) +
  #scale_colour_viridis_d(end = 0.9) +
  theme_dark(base_size = 18) +
  labs(title = "Flowering period from 1945-2012 at 7 sites", subtitle = "median start day of year to median end day of year", caption = "1500 forcing observations simulated  from 200 draws of the posterior with new factor levels \n and matched to forcing data for plotted sites and years. Daily temperature data from PCIC \nand adjusted using monthly climateNA") +
  theme(legend.position = "none") +
  scale_y_date(date_labels = "%b %e") +
  scale_colour_viridis_d()

# widemedians <- general_doy_preds_med_siteyearsex %>% 
#   filter(.width == 0.5) %>%
#   select(Sex, event, Site, Year, newdoycol) %>% 
#   pivot_wider(names_from = event, values_from = newdoycol)


```
## Historical overlap

30 forcing samples from the model translated into DoY of flowering event for 7 Sites 1945-2012. Then calculated median DoY across samples within each site, year, and sex combination and used those to construct flowering period intervals (begin to end). Then determined the intersection of those intervals for all sites and years

```{r calculate overlap}

med_sys_begin <- filter(general_doy_preds_med_siteyearsex, event == "begin") %>%
  rename(b.newdoycol = newdoycol, b.lower = .lower, b.upper = .upper, b.date = Date, b.lowerdate = .lowerdate, b.upperdate = .upperdate) %>%
  select(-event)
med_sys_end <- filter(general_doy_preds_med_siteyearsex, event == "end") %>%
   rename(e.newdoycol = newdoycol, e.lower = .lower, e.upper = .upper, e.date = Date, e.lowerdate = .lowerdate, e.upperdate = .upperdate) %>%
  select(-event)

med_interval_sys <- full_join(med_sys_begin, med_sys_end) %>%
  mutate(med_interval = lubridate::interval(b.date, e.date, tzone = "UTC")) %>% # calculate interval for phenology
  select(Sex, Site, Year, .width, .point, .interval, med_interval) 

female_intervals <- filter(med_interval_sys, Sex == "FEMALE") %>%
  rename(female_med_interval = med_interval) %>%
  select(-Sex)
  
male_intervals <- filter(med_interval_sys, Sex == "MALE") %>%
  rename(male_med_interval = med_interval, male_Site = Site) %>%
  select(-Sex)

med_overlap_sys <- full_join(female_intervals, male_intervals) %>%
  mutate(overlap = day(as.period(intersect(female_med_interval, male_med_interval), "days")) + 1, Site = forcats::fct_relevel(Site, sitefactororder), male_Site = forcats::fct_relevel(male_Site, sitefactororder)) 
med_overlap_sys[is.na(med_overlap_sys)] <- 0

ggplot(filter(med_overlap_sys, .width == 0.5), aes(x = 1, y = overlap, fill = Site )) +
  geom_violin(draw_quantiles = c(0.5), alpha = 0.5) +
  facet_grid(Site ~ male_Site) +
  scale_fill_viridis_d(option = "cividis") +
  labs(title = "Days of flowering overlap between sites", subtitle = "1945-2012", caption = "30 forcing samples from the model translated into DoY of flowering event for 7 Sites 1945-2012. Then calculated median DoY across samples and used those to construct flowering period intervals (begin to end). Then determined the intersection of those intervals for all sites and years")
```



# Climate change

Daily mean temperatures for 2011-2100 were calculated for all sites. Mean monthly temperatures from a 13 model ensemble developed by @Mahony2021 were used to adjust 6 years of daily temperatures from the historical record up or down. I used every 15 years between 1945 and 2012 to cover a range of time periods and yearly variability. Mean monthly temperatures for each past year were compared to mean monthly temperatures for each future year. The difference between MMT future year and historical year was used to adjust daily temperatures for each historical time series so the mmt is the same for each.

```{r future, echo = FALSE}
# calculate forcing for future clim
futclimf <- futclim %>%
  # calculate heatsum forcing with 5 degree threshold
  mutate(forcing = case_when(mean_temp_gcm >= 5 ~ mean_temp_gcm - 5,
                             mean_temp_gcm < 5 ~ 0) )%>%
  arrange(SSP, Site, Year, normal_period, DoY) %>%
  group_by(SSP, Site, Year, normal_period) %>%
  mutate(sum_forcing = cumsum(forcing)) %>%
  ungroup() %>% group_by(Site, Year, name) %>%
  group_split()


# split this dataframe (gorup_split) by Site, Year, and name and then use purrr map to run find day of forcing function
# DoY predictions for future (general) ####
 
# select only columns necessary for matching to DoY in climate dataset
smallgs <- generalsim %>% 
  ungroup() %>%
  select(.prediction)%>% 
  arrange(.prediction)



doypredmatchfut <- purrr::map(futclimf, match_force_future, bdf = smallgs, aforce = "sum_forcing", bforce = "prediction")  %>%
  bind_rows() %>%
  full_join(generalsim)

doypredmatchfut_medians <- doypredmatchfut %>% group_by(Sex, event, Site, SSP, normal_period, climate_forcing) %>%
  point_interval(newdoycol, .width = c(0.5, 0.95), .point = median, .interval = hdci) %>%
  mutate(Date = as.Date("2020-12-31") + newdoycol, .lowerdate = as.Date("2020-12-31") + .lower, .upperdate = as.Date("2020-12-31") + .upper)

ggplot(filter(doypredmatchfut_medians, climate_forcing %in% c(4.5, 8.5)), aes(y = Date, x = normal_period, ymin = .lowerdate, ymax = .upperdate, group = interaction(normal_period, Sex), colour = Sex)) +
  geom_pointinterval(position = "dodge", alpha = 0.5)  +
  facet_grid(climate_forcing ~ Site) +
  #scale_colour_viridis_d(end = 0.9) +
  theme_dark(base_size = 18) +
  labs(title = "Future flowering periods at 7 sites for 2 Climate forcing scenarios", subtitle = "median start day to median end day", caption = "medians of 1500 forcing observations simulated from 30 draws of the posterior with new factor levels and matched \nto day of year data for plotted sites and years. Daily temperature timeseries for 7 sites from PCIC & adjusted using ClimateNA") +
  #theme(legend.position = "none") +
  scale_colour_viridis_d() +
  scale_y_date(date_labels = "%b %e") +
  theme(axis.text.x = element_text(angle = 30, hjust=1), legend.position = "top") +
  xlab("Normal period")
```
Date of flowering is earlier as climate change progresses - by quite a lot under some climate change scenarios.

# length
```{r length}

#individual lengths
specific_doy_preds_length <- specific_doy_preds %>%
  ungroup() %>%
  select(-prediction, -sum_forcing, -upper, -contains("censored")) %>%
  pivot_wider(names_from = event, values_from = newdoycol) %>%
  mutate(length = end - begin) 

ggplot(specific_doy_preds_length, aes(x = length, y = prediction_type, colour = Sex)) +
  stat_histinterval(position = "dodge", .width = c(0.5, 0.89)) +
  scale_colour_viridis_d() +
  labs(title = "Length of phenological period for individuals")

# population lengths
specific_doy_preds_length_ts <- specific_doy_preds %>%
  group_by(Sex, event, Site, Year, prediction_type) %>%
  median_hdci(newdoycol, .width = c(0.5, 0.89)) %>%
  # drop uncertainty for now
  filter(.width == 0.5) %>%
  select(-starts_with(".")) %>% distinct() %>%
  pivot_wider(names_from = event, values_from = newdoycol) %>%
  mutate(length = end - begin) %>%
  ungroup() %>%
  mutate(Site = forcats::fct_relevel(as.factor(Site), sitefactororder))


ggplot(filter(specific_doy_preds_length_ts, prediction_type == "prediction - full cross"), aes(x = Site, y = length, colour = Sex)) +
  geom_beeswarm(dodge.width =0.75) +
  scale_colour_viridis_d() +
  labs(title = "Median flowering period length at a site", subtitle = "Each point represents one year at one site 1997-2012", caption = "fully crossed predictions") +
  theme(legend.position = "bottom") +
  ylab("Length of flowering period (days)") 

futlen <- doypredmatchfut_medians %>% 
  select(-contains("Date"), -contains(".lower"), -contains(".upper"), -.width) %>%
  distinct() %>%
  tidyr::pivot_wider(names_from = event, values_from = newdoycol) %>%
  mutate(period_length = end - begin) 

ggplot(filter(futlen, climate_forcing %in% c(4.5, 8.5)), aes(y = period_length, x = normal_period, colour = Sex)) +
  geom_point()  +
  facet_grid(climate_forcing ~ Site) +
  #scale_colour_viridis_d(end = 0.9) +
  theme_dark(base_size = 18) +
  #theme(legend.position = "none") +
  scale_colour_viridis_d() +
  theme(legend.position = "bottom") +
  ylab("Length of flowering period (days)") +
  labs(title = "Future flowering period median length", caption = "fully crossed - only 5 posterior samples!") +
  theme(axis.text.x = element_text(angle = 30, hjust=1), legend.position = "top") +
  xlab("Normal period") 

# histlen_uncertainty <- general_doy_preds_med_siteyearsex %>%
#   group_by(event) %>%
#   pivot_wider(values_from)
# 
# histlen <- general_doy_preds_med_siteyearsex %>%
#   select(-starts_with(".")) %>%
#   distinct() %>%
#   tidyr::pivot_wider(names_from = event, values_from = newdoycol) %>%
#   mutate(period_length = end - begin)
# 
# ggplot(histlen, aes(x = Year, y = period_length, linetype = Sex)) +
#   geom_line() +
#   facet_grid(Site ~ .)
# 
# ggplot(histlen, aes(x = Site,  y = period_length, colour = Sex)) +
#   geom_quasirandom(dodge.width = 0.75) +
#   scale_colour_viridis_d() 
# 
# 
# ggplot(futlen, aes(x = as.numeric(as.factor(normals)), y = period_length, colour = climate_forcing, linetype = Sex, shape = as.factor(climate_forcing))) +
#   geom_point(alpha = 0.8) +
#   geom_line(alpha = 0.9) +
#   facet_grid(. ~ Site) +
#   scale_colour_viridis_c() +
#   theme(legend.position = "top")

```
Haven't figured out a good graph for future length yet. Not sure it's really doing anything?

# variation
```{r doyvariation}
doyvar <- specific_doy_preds %>%
  group_by(Sex, event, Site, Year, Provenance, Clone, Tree) %>%
  summarise(sd = sd(newdoycol)) %>%
  median_hdci(sd, .width = c( 0.5, 0.89)) 
  
```

# overlap

```{r overlap}
#specific_doy_preds %>% filter(prediction_type == "retrodiction - uncensored")
```
